Search for Spatially Extended \fermi-LAT Sources Using Two Years of Flight Data

### 2011-08-28 extended_catalog/paper/extended_catalog.tex
renaming
josh revisions

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\newcommand{\km}{\text{km}\xspace}
\newcommand{\phflux}{\ensuremath{\ph\ \cm^{-2}\s^{-1}}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}

@@

@@

The current generation of air Cherenkov detectors have
made it apparent that many sources can be spatially resolved
at even higher energies.
Most
prominent was a survey of the Galactic plane using the High Energy
Stereoscopic System (H.E.S.S) which reported 14 spatially extended
sources with extensions varying from $\sim0\fdg1$ to $\sim0\fdg25$
\citep{hess_plane_survey}.  In fact, within our Galaxy only very few
sources detected at \tev energies (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61$-$303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes})
have no detectable extension.  High-energy $\gamma$-rays from
these sources are produced by the decay of $\pi^0$s produced by hadronic
interactions with interstellar matter and by relativistic electrons
due to Inverse Compton (IC) scattering and Bremsstrahlung radiation
\citep{blandford_and_eichler_1987}.  It is likely that the \gev and
\tev emission from these sources originates from the same population of
high-energy particles and so at least some of these sources should be
detected at \gev energies.  Studying 
these \tev sources at \gev energies would help to
determine the emission mechanisms producing these high energy photons.

The LAT is a pair conversion telescope that has been surveying the
$\gamma$-ray sky since 2008 June.  The LAT has broad energy coverage
(20 \mev to $>300$ \gev), wide field of view ($\sim 2.4 \sr$), and large
effective area ($\sim 8000\ \cm^2$ at $>1 \gev$).  
Using one year of all-sky surveying data, the LAT Collaboration published
a catalog of 1451 sources significantly detected at \gev energies called 1FGL \citep{first_cat}.
Using two years of data, a second catalog called 2FGL reported on 1873
sources \citep{second_cat}.
The counterparts of many of these sources can be spatially resolved
when observed at other frequencies but detecting the spatial extension
of these sources at \gev energies is difficult because the size of the
point-spread function (PSF) of the LAT is comparable to the typical size
of many of these sources.

@@

Previous analysis of
extended LAT sources were performed as dedicated studies
of individual sources, so we expect
that a systematic scan of all LAT detected sources
could uncover additional spatially extended sources.  For these
reasons, in Section~\ref{analysis_methods_section} we present
a new systematic method for analyzing spatially extended LAT sources. In
Section~\ref{monte_carlo_validation} we demonstrate that this method
can be used to test the statistical significance of
the extension of
a LAT sources and in Section~\ref{extension_sensitivity} we calculate
the LAT's detection threshold to resolve the extension of a source.
In Section~\ref{extended_source_search_method} we describe a search for
new spatially extended LAT sources and in Section~\ref{validate_known}
we reanalyze the twelve previously detected extended sources
included in 2FGL.  In Section~\ref{test_2lac_sources} we further
demonstrate that our detection method does not misidentify
point sources as being extended 
by testing the extension of active
galactic nuclei (AGN) believed to be unresolvable. Finally, in
Section~\ref{new_ext_srcs_section} we present the detection of the
extension of nine spatially extended sources that were reported in 2FGL
but not previously spatially resolved.

@@

Morphological studies of sources using the LAT are challenging
because of the strongly energy-dependent PSF that is comparable in
size to the extension of many sources expected to be detected at
\gev energies.  Additional complications arise for sources along
the Galactic plane due to systematic uncertainties in the Galactic
diffuse emission.  The LAT's PSF is limited at its lowest detectable
energies by multiple scattering in the silicon strip tracking section
of the detector and is several degrees at 100 \mev.  The PSF improves
with energy approaching a 68\% containment radius of $\sim0\fdg2$ at
the highest energies (when averaged over the acceptance of the LAT)
and is limited by the granularity of the silicon strips in the tracker
\citep{atwood_LAT_mission,on_orbit_calibration,lat_on_orbit_psf}.\footnote{More
information about the performance of the LAT can be found at the \fermi
Science Support Center (FSSC,\url{http://fermi.gsfc.nasa.gov}).} However,
since most high energy astrophysical sources have spectra that decrease
rapidly with increasing energy, the improved resolution of the higher
energy photons is offset by their low statistics. Therefore sophisticated
analysis techniques are required to maximize our sensitivity
to these extended sources.

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool performs
a maximum likelihood analysis in which the Poisson likelihood to find
the observing counts is maximized given a parametrized spatial and
spectral model of the source and its surrounding region.  The sky is
divided into cubes in space and energy using the healpix representation
of the sky \citep{healpix} and the likelihood is maximized over all bins
in a region.  The extension of a source can be modeled by a geometric
shape (e.g. a disk or a two-dimensional Gaussian) and the source's position, extension,
and spectrum can be simultaneously fit.

@@

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT Collaboration called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  The choice to base the
spatial extension fitting on \pointlike rather than \gtlike was made
on considerations of computing time.  The \pointlike algorithm was
optimized for speed to handle larger numbers of sources efficiently
which is important for our catalog scan and for being able
to perform large-scale Monte Carlo simulations to validate the tool.
Details on the \pointlike package can be
found in \cite{matthew_kerr_thesis}.  We extended the code to allow a
simultaneous fit of the source extension together with the position and
the spectral parameters.

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial
distribution and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library 
\citep{minuit_documentation}
to maximize the
likelihood by simultaneously varying the position and extension of
the source.  As will be described in
Section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is important to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

The significance of the extension of a source can be calculated from the
likelihood ratio test of a model with a spatially extended source and
a model with a point-like source. The test statistic for this procedure
is defined as
\begin{equation}
  \tsext=2\log(\likelihood_\text{ext}/\likelihood_\text{ps}) 
\end{equation}
where \likelihood is the Poisson likelihood.
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
Section~\ref{monte_carlo_validation}.

For extended sources with an assumed radially symmetric shape,
we can optimize the calculation by performing one
of the integrals analytically.
The expected photon 
distribution can be written as
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') r' dr' d\phi'
\end{equation}
For the LAT, the PSF can be parameterized by a King function \citep{king_function}:
\begin{equation}
  \text{PSF}(r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $u=(r/\sigma)^2/2$ and $\sigma$ and $\gamma$ are free parameters
\citep{matthew_kerr_thesis}.  For radially symmetric extended sources,
the angular part of the integral can be evaluated analytically
\begin{align}
  \text{PDF}(u) & = \int_0^\infty r' dr'
  I_\text{src}(v) 
  \int_0^{2\pi} d\phi' 
  \text{PSF}(\sqrt{2\sigma^2(u+v-2\sqrt{uv}\cos(\phi-\phi'))})
  \\
  & = \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times {}_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{align}
where $v=(r'/\sigma)^2/2$ and ${}_2F_1$ is the Gaussian hypergeometric
function.  This convolution formula reduces the expected photon
distribution to a single numerical integral.

There will always be a small numerical discrepancy between the expected
photon distribution derived from a true point source and a very small
extended source due to numerical error in the convolution.  In most
situations, this error is insignificant.  But in particular for
very bright sources, this numerical error has the potential to bias the
test statistic for the extension test. Therefore, when calculating
\tsext, we compare the likelihood fitting the source with an extended
spatial model to the likelihood when fixing its extension to $10^{-10\,\circ}$.

We estimate the error on the extension of a source by fixing
the position of the source and varying the extension until the
likelihood has fallen by \onehalf, corresponding to a $1\sigma$ error
\citep{Statistical_methods_book}.  This method is shown schematically in
Figure~\ref{four_plots_ic443} which shows the change in the log of the
likelihood when varying the extension of the SNR IC443.  The localization
error is calculated by fixing the extension and spectrum of the source
and fitting to the likelihood function a 2D Gaussian as a function
of position.

@@

There is a degeneracy between a spatially extended source and multiple
point sources separated by angular distances comparable to or smaller than
the size of the LAT PSF.  To assess the possibility of source confusion,
we use \pointlike to simultaneously fit the position of two point sources
in the region of extended source candidates.

We define \tsinc as twice the increase in the log of the likelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\log(\likelihood_\text{2pts}/\likelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Similar to the case of extended sources described in Section~\ref{gtlike_crosscheck},
the spectra of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We tested two different models for the
surface brightness profile, either a 2D Gaussian
\begin{equation}\label{gauss_pdf}
  I(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}\label{disk_pdf}
  I(x,y)=
  \begin{cases}
    \frac{1}{\pi\sigma^2} & x^2+y^2\le\sigma^2 \\
    0                      & x^2+y^2>\sigma^2
  \end{cases}
\end{equation}
Although these shapes are significantly different, when convolved
with the PSF, their profiles are very similar.  
This is demonstrated in 
Figure~\ref{compare_disk_gauss}
by showing $I_\text{src}$, the PSF, and the PDF for
a radially symmetric uniform brightness and a Gaussian surface brightness of size 0\fdg5.\footnote{To
allow a valid comparison between Gaussian and disk shaped morphologies,
we define the source size for this test as the radius containing
68\% of the intensity ($\rsixeight$).  $\rsixeight_\text{,Gaussian}=1.51\sigma$
and $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in Equation~\ref{gauss_pdf} and
Equation~\ref{disk_pdf} respectively.} This plot shows that the LAT 
has little sensitivity to
the exact structure of an extended source.
Therefore, in our
search for extended sources, we use only a uniform
disk as our spatial extension shape. We then quote the radius of the
disk edge as the size of the source.

@@

We test the false detection probability of \pointlike by fitting the extension of
point-like sources.
\cite{mattox_egret} discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is 
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\onehalf(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of Equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension since the same conditions
apply (with the source flux $\Phi$ replaced by the source radius $r$ and
$r<0$ being unphysical).
To validate this claim, we performed Monte Carlo
simulations
to calculate empirical distributions for $\tsext$ and
compared them to Equation~\ref{ts_ext_distribution}.

We simulated point sources with various spectral forms using
the LAT on-orbit simulation tool, 
\gtobssim\footnote{\gtobssim is distributed publicly by the FSSC.} and fit them with \pointlike using both point
and extended source hypotheses.  These point sources were simulated with a power-law
spectral model with integrated fluxes above 100 \mev ranging from $3\times10^{-9}$ 
to $1\times10^{-6}$ \phflux in six discrete steps and spectral
indices ranging from 1.5 to 3 in four discrete steps.  These values
were picked to to represent typical source parameters of LAT-detected
sources. The point sources were simulated on top of an isotropic
background with an integrated flux above 100 \mev of $1.5\times10^{-5}$ \phflux
taken to be the same as the isotropic spectrum measured by EGRET
\citep{sreekumar_isotropic}.  The Monte Carlo simulation was performed
over a one-year observation period using a representative rocking profile and a
representative livetime fraction of 0.8.  The reconstruction was performed
using 1 \gev to 100 \gev photons and the Pass~7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs, \cite{lat_on_orbit_psf}).  For each 
significantly detected point source ($\ts\ge25$), we used \pointlike
to fit it as an extended source and calculate \tsext.

For each set of spectral parameters, $\sim30,000$ statistically independent
simulations were performed. For the dimmer spectral models, many of the
simulations left the source undetected ($\ts<25$)
and were discarded.  Table~\ref{ts_ext_num_sims}
shows the different spectral models used in our study as well as the
number of simulations.  The cumulative density of \tsext is plotted in
Figure~\ref{ts_ext_mc}. The $\chi^2_1/2$ distribution of
Equation~\ref{ts_ext_distribution} is overlaid for comparison.

Our study shows broad agreement between simulations and
Equation~\ref{ts_ext_distribution}. Nevertheless, the agreement is not
perfect.  It should be noted that the discrepancy seems to be worst for
bright sources where numerical errors in the convolution
are most apparent.  Another possible
reason for the departure from Equation~\ref{ts_ext_distribution} 
is that \pointlike ignores energy dispersion which will change the
PSF shape as a function of energy. We emphasize that many of
the empirical distributions lie to the left of the theoretical curve so
using the theoretical distribution will lead to an underestimate of the
statistical significance of a detection. Therefore, we are confident that
$\sqrt{\tsext}$ can be used as a conservative measure of the statistical
significance of a source's extension and use it in the following analysis.

@@

We calculated the LAT's detection threshold to detect that a spatially extended
sources is extended. We define the detection threshold as the flux at which the value
of $\tsext$ averaged over many statistical realizations of a source
is $\langle\tsext\rangle=16$, corresponding to a $4\sigma$ detection
(see section~\ref{monte_carlo_validation}).

We used a simulation setup similar to that described in
Section~\ref{monte_carlo_validation}, but instead of point sources
we simulated extended
sources with a radially symmetric uniform surface
brightness. Additionally, we
simulated our sources over the two-year time range included
in 2FGL.
Each extension and
spectral index, we selected a flux range which bracketed $\tsext=16$
and performed an extension test for $>100$ independent realizations of
ten flux values in this range.
We calculated $\langle\tsext\rangle=16$ by fitting a line to the flux
and $\tsext$ values in this narrow range.

Figure~\ref{index_sensitivity} shows the threshold for sources of four
spectral indices from 1.5 to 3 and extension varying from $\sigma=0\fdg1$
to $2\fdg0$.  The LAT's flux threshold for a significant detection
of source extension drops quickly with
increasing source size and reaches a minimum around 0\fdg5. 
Figure~\ref{index_sensitivity} shows
the threshold using photons with energies between 100 \mev and 100 \gev
and also using only 1 \gev to 100 \gev photons.
Except for very large ($>1\deg$) sources, our detection threshold is
not substantially improved by including photons with energies between 100 \mev and
1 \gev.  This is also demonstrated in Figure~\ref{four_plots_ic443}
which shows \tsext for the SNR IC443 computed independently in twelve
energy bins between 100 \mev and 100 \gev. For IC443, which has a
spectral index $\sim2.4$, almost the entire increase in likelihood
modeling the source as being extended comes
from energies above 1 \gev.  On the other hand, other systematic errors
become increasingly important at low energy. For our extension search,
we therefore use only photons with energies above 1 \gev.

>>>>

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\newcommand{\km}{\text{km}\xspace}
\newcommand{\phflux}{\ensuremath{\text{ph}\ \text{cm}^{-2}\,\text{s}^{-1}}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}

@@

\newcommand{\degree}{^\circ\xspace}

@@

The current generation of air Cherenkov detectors have
made it apparent that many sources can be spatially resolved
at even higher energies.
Most
prominent was a survey of the Galactic plane using the High Energy
Stereoscopic System (H.E.S.S) which reported 14 spatially extended
sources with extensions varying from $\sim0\fdg1$ to $\sim0\fdg25$
\citep{hess_plane_survey}.  In fact, within our Galaxy only very few
sources detected at \tev energies (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61$-$303 \citep{MAGICLSI, VERITASLSI},
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes})
have no detectable extension.  High-energy $\gamma$-rays from
these sources originate by the decay of $\pi^0$s produced by hadronic
interactions with interstellar matter and by relativistic electrons
due to Inverse Compton (IC) scattering and Bremsstrahlung radiation
\citep{blandford_and_eichler_1987}.  It is likely that the \gev and
\tev emission from these sources originates from the same population of
high-energy particles and so at least some of these sources should be
detected at \gev energies.  Studying 
these \tev sources at \gev energies would help to
determine the emission mechanisms producing these high energy photons.

The LAT is a pair conversion telescope that has been surveying the
$\gamma$-ray sky since 2008 June.  The LAT has broad energy coverage
(20 \mev to $>300$ \gev), wide field of view ($\sim 2.4$ sr), and large
effective area ($\sim 8000\ \cm^2$ at $>1 \gev$).  
Using one year of all-sky surveying data, the LAT Collaboration published
a catalog of 1451 sources significantly detected at \gev energies called 1FGL \citep{first_cat}.
Using two years of data, a second catalog called 2FGL reported on 1873
sources \citep{second_cat}.
The counterparts of many of these sources can be spatially resolved
when observed at other frequencies but detecting the spatial extension
of these sources at \gev energies is difficult because the size of the
point-spread function (PSF) of the LAT is comparable to the typical size
of many of these sources.

@@

Previous analysis of
extended LAT sources were performed as dedicated studies
of individual sources, so we expect
that a systematic scan of all LAT detected sources
could uncover additional spatially extended sources.  For these
reasons, in Section~\ref{analysis_methods_section} we present
a new systematic method for analyzing spatially extended LAT sources. In
Section~\ref{monte_carlo_validation}, we demonstrate that this method
can be used to test the statistical significance of
the extension of
a LAT sources and in Section~\ref{extension_sensitivity} we calculate
the LAT's detection threshold to resolve the extension of a source.
In Section~\ref{extended_source_search_method}, we describe a search for
new spatially extended LAT sources and in Section~\ref{validate_known}
we reanalyze the twelve previously detected extended sources
included in 2FGL.  In Section~\ref{test_2lac_sources} we further
demonstrate that our detection method does not misidentify
point-like sources as being extended 
by testing the extension of active
galactic nuclei (AGN) believed to be unresolvable. Finally, in
Section~\ref{new_ext_srcs_section} we present the detection of the
extension of nine spatially extended sources that were reported in 2FGL
but not previously spatially resolved.

@@

Morphological studies of sources using the LAT are challenging
because of the strongly energy-dependent PSF that is comparable in
size to the extension of many sources expected to be detected at
\gev energies.  Additional complications arise for sources along
the Galactic plane due to systematic uncertainties in the Galactic
diffuse emission.  The LAT's PSF is limited at its lowest detectable
energies by multiple scattering in the silicon strip tracking section
of the detector and is several degrees at 100 \mev.  The PSF improves
with energy approaching a 68\% containment radius of $\sim0\fdg2$ at
the highest energies (when averaged over the acceptance of the LAT)
and is limited by the granularity of the silicon strips in the tracker
\citep{atwood_LAT_mission,on_orbit_calibration,lat_on_orbit_psf}.\footnote{More
information about the performance of the LAT can be found at the \fermi
Science Support Center (FSSC, \url{http://fermi.gsfc.nasa.gov}).} However,
since most high energy astrophysical sources have spectra that decrease
rapidly with increasing energy, the improved resolution of the higher
energy photons is typically offset by their low statistics. Therefore sophisticated
analysis techniques are required to maximize the LATs sensitivity
to these extended sources.

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT.
The Poisson likelihood to find
the observing counts is maximized given a parametrized spatial and
spectral model of the source and its surrounding region.  The sky is
divided into cubes of space and energy using the healpix representation
of the sky \citep{healpix} and the likelihood is maximized over all bins
in a region.  The extension of a source can be modeled by a geometric
shape (e.g. a disk or a two-dimensional Gaussian) and the source's position, extension,
and spectrum can be simultaneously fit.

@@

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT Collaboration called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  The choice to base the
spatial extension fitting on \pointlike rather than \gtlike was made
on considerations of computing time.  The \pointlike algorithm was
optimized for speed to handle larger numbers of sources efficiently
which is important for our catalog scan and for being able
to perform large-scale Monte Carlo simulations to validate the analysis.
Details on the \pointlike package can be
found in \cite{matthew_kerr_thesis}.  We extended the code to allow a
simultaneous fit of the source extension together with the position and
the spectral parameters.

@@

In \pointlike, it is assumed that the spatial and spectral model
of an extended source are separable, i.e. that the source model
$M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial distribution
and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a function
of energy) and uses the \minuit library \citep{minuit_documentation}
to maximize the likelihood by simultaneously varying the position,
extension, and spectrum of the source.  As will be described in
Section~\ref{monte_carlo_validation}, simultaneously fitting the
position, extension, and spectrum is important to correctly calculate
the statistical significance of the detection of extension of a source.
To avoid projection effects, the source's longitude and latitude are
not directly fit but instead the source's displacement in a rotated
reference frame.

The significance of the extension of a source can be calculated from the
likelihood ratio test of a hypothesis with a spatially extended source and
a hypothesis with a point-like source. The test statistic for this procedure
is defined as
\begin{equation}
  \tsext=2\log(\likelihood_\text{ext}/\likelihood_\text{ps}) 
\end{equation}
where \likelihood is the Poisson likelihood.
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point-like source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
Section~\ref{monte_carlo_validation}.

For extended sources with an assumed radially symmetric shape,
we optimized the calculation by performing one
of the integrals analytically.
The expected photon 
distribution can be written as
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') r' dr' d\phi'
\end{equation}
For the LAT, the PSF can be parameterized by a King function \citep{king_function}:
\begin{equation}
  \text{PSF}(r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $u=(r/\sigma)^2/2$ and $\sigma$ and $\gamma$ are free parameters
\citep{matthew_kerr_thesis}.  For radially symmetric extended sources,
the angular part of the integral can be evaluated analytically
\begin{align}
  \text{PDF}(u) & = \int_0^\infty r' dr'
  I_\text{src}(v) 
  \int_0^{2\pi} d\phi' 
  \text{PSF}(\sqrt{2\sigma^2(u+v-2\sqrt{uv}\cos(\phi-\phi'))})
  \\
  & = \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times {}_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{align}
where $v=(r'/\sigma)^2/2$ and ${}_2F_1$ is the Gaussian hypergeometric
function.  This convolution formula reduces the expected photon
distribution to a single numerical integral.

There will always be a small numerical discrepancy between the expected
photon distribution derived from a true point-like source and a very small
extended source due to numerical error in the convolution.  In most
situations, this error is insignificant.  But in particular for
very bright sources, this numerical error has the potential to bias the
test statistic for the extension test. Therefore, when calculating
\tsext, we compare the likelihood fitting the source with an extended
spatial model to the likelihood when fixing its extension to $10^{-10\,\circ}$.

We estimate the error on the extension of a source by fixing
the position of the source and varying the extension until the
likelihood has fallen by \onehalf, corresponding to a $1\sigma$ error
\citep{Statistical_methods_book}.  Figure~\ref{four_plots_ic443}
demonstrates this method by showing the change in the log of the
likelihood when varying the extension of the SNR IC443.  The localization
error is calculated by fixing the extension and spectrum of the source
and fitting to the likelihood function a 2D Gaussian as a function
of position.

@@

There is a degeneracy between a spatially extended source and multiple
point-like  sources separated by angular distances comparable to or smaller than
the size of the LAT PSF.  To assess the possibility of source confusion,
we use \pointlike to simultaneously fit the position of two point-like sources
in the region of extended source candidates.

We define \tsinc as twice the increase in the log of the likelihood fitting the
region as two point-like sources compared to fitting the region as one point-like
source: 
\begin{equation}
  \tsinc=2\log(\likelihood_\text{2pts}/\likelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
and $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Similar to the case of extended sources described in Section~\ref{gtlike_crosscheck},
the spectra of the two point-like sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We tested two different models for the
surface brightness profile, either a 2D Gaussian
\begin{equation}\label{gauss_pdf}
  I(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform surface brightness
\begin{equation}\label{disk_pdf}
  I(x,y)=
  \begin{cases}
    \frac{1}{\pi\sigma^2} & x^2+y^2\le\sigma^2 \\
    0                      & x^2+y^2>\sigma^2.
  \end{cases}
\end{equation}
Although these shapes are significantly different, 
Figure~\ref{compare_disk_gauss} shows the their PDF 
are very similar for an example source of size
0\fdg5.\footnote{To
allow a valid comparison between the Gaussian and the uniform surface
brightness morphologies,
we define here the source size (0\fdg5) 
as the radius containing
68\% of the intensity ($\rsixeight$).  $\rsixeight_\text{,Gaussian}=1.51\sigma$
and $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in Equation~\ref{gauss_pdf} and
Equation~\ref{disk_pdf} respectively.} The LAT 
has little sensitivity /thereto the exact structure of extended source.
Therefore, in our search for extended sources we used only a uniform
surface brightness as our spatial extension shape. We quote the radius
of the edge ($\sigma$) as the size of the source.

@@

We test the false detection probability of for the
extended source test by fitting the extension of
point-like sources.
\cite{mattox_egret} discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is 
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\onehalf(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of Equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom
and leads to the often quoted result that the square root
of the test statistic is the number of $\sigma$ of
the detection. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension since the same conditions
apply (with the source flux $\Phi$ replaced by the source radius $r$ and
$r<0$ being unphysical).
We performed Monte Carlo
simulations
to calculate empirical distributions for $\tsext$ and
we compared them to Equation~\ref{ts_ext_distribution}.

We simulated point-like sources with various spectral forms using
the LAT on-orbit simulation tool
\gtobssim\footnote{\gtobssim is distributed publicly by the FSSC.} and fit the sources
with \pointlike using both point-like
and extended source hypotheses.  These point-like sources were simulated with a power-law
spectral model with integrated fluxes above 100 \mev ranging from $3\times10^{-9}$ 
to $1\times10^{-6}$ \phflux in six discrete steps and spectral
indices ranging from 1.5 to 3 in four discrete steps.  These values
were picked to represent typical source parameters of LAT-detected
sources. The point-like sources were simulated on top of an isotropic
background with an integrated flux above 100 \mev of $1.5\times10^{-5}$ \phflux
taken to be the same as the isotropic spectrum measured by EGRET
\citep{sreekumar_isotropic}.  The Monte Carlo simulation was performed
over a one-year observation period using a representative rocking profile and a
representative livetime fraction of 0.8.  The reconstruction was performed
using photons with energies between 1 \gev to 100 \gev 
and the Pass~7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs, \cite{lat_on_orbit_psf}).  For each 
significantly detected point-like source ($\ts\ge25$), we used \pointlike
to fit the source as an extended source and calculate \tsext.

For each set of spectral parameters, $\sim30,000$ statistically independent
simulations were performed. For dimmer spectral models, many of the
simulations left the source undetected ($\ts<25$)
and were discarded.  Table~\ref{ts_ext_num_sims}
shows the different spectral models used in our study as well as the
number of simulations and the average point-like source
significance.  The cumulative density of \tsext is plotted in
Figure~\ref{ts_ext_mc} and compared to the $\chi^2_1/2$ distribution of
Equation~\ref{ts_ext_distribution}.

Our study shows broad agreement between simulations and
Equation~\ref{ts_ext_distribution}. Nevertheless, the agreement is not
perfect.  It should be noted that the discrepancy seems to be worst for
bright sources where numerical errors in the convolution
are most apparent.  Another possible
reason for the departure from Equation~\ref{ts_ext_distribution} 
is that \pointlike ignores energy dispersion which will change the
PSF shape as a function of energy. We emphasize that many of
the empirical distributions lie to the left of the theoretical curve so
using the theoretical distribution will often lead to an underestimate of the
statistical significance of a detection. Therefore, we are confident that
\tsext can be used as a conservative measure of the statistical
significance of a source's extension and we used it in the following analysis.

@@

We calculated the LAT's threshold to detect that a spatially extended
source is extended. We define the detection threshold as the flux at which the value
of $\tsext$ averaged over many statistical realizations of a source
is $\langle\tsext\rangle=16$, corresponding to a $4\sigma$ detection.

We used a simulation setup similar to that described in
Section~\ref{monte_carlo_validation}, but instead of point-like sources
we simulated extended
sources with a radially symmetric uniform surface
brightness. Additionally, we
simulated our sources over the two-year time range included
in 2FGL.
For each extension and
spectral index, we selected a flux range which bracketed $\tsext=16$
and performed an extension test for $>100$ independent realizations of
ten fluxes in the range.
We calculated $\langle\tsext\rangle=16$ by fitting a line to the flux
and $\tsext$ values in the narrow range.

Figure~\ref{index_sensitivity} shows the threshold for sources of four
spectral indices from 1.5 to 3 and extensions varying from $\sigma=0\fdg1$
to $2\fdg0$.  The LAT's flux threshold for a significant detection
of source extension drops quickly with
increasing source size and reaches a minimum around 0\fdg5. 
Figure~\ref{index_sensitivity} shows
the threshold using photons with energies between 100 \mev and 100 \gev
and also using only 1 \gev to 100 \gev photons.
Except for very large ($>1\degree$) sources, our detection threshold is
not substantially improved by including photons with energies between 100 \mev and
1 \gev.  This is also demonstrated in Figure~\ref{four_plots_ic443}
which shows \tsext for the SNR IC443 computed independently in twelve
energy bins between 100 \mev and 100 \gev. For IC443, which has a
spectral index $\sim2.4$, almost the entire increase in likelihood
modeling the source as being extended comes
from energies above 1 \gev.  On the other hand, other systematic errors
become increasingly important at low energy. For our extension search,
we therefore used only photons with energies above 1 \gev.

\hl{
Figure~\ref{all_sensitivity} shows the flux threshold as a function of
source extension for different background levels ($1\times$, $10\times$,
and $100\times$ the nominal background), different spectral indices,
and two different energy ranges (1 \gev to 100 \gev and 10 \gev to
100 \gev).  The detection threshold is higher for sources in regions of
higher background.  When studying sources only at energies above 1 \gev,
the LAT's detection threshold (defined as the 1 \gev to 100 \gev flux at
which $\langle\tsext\rangle=16$) depends less strongly on the source's
spectral index. The index dependence of the detection threshold is
even weaker when considering only photons with energies above 10 \gev.
Overlaid on Figure~\ref{all_sensitivity} are the LAT detected extended
sources that will be discussed in Sections~\ref{validate_known} and
\ref{new_ext_srcs_section}.  The extension thresholds are tabulated in
Table~\ref{all_sensitivity_table}.
